\textbf{ Interpretable Gaussian Processes for Stellar Variability }

### 2021-01-12 tex/ms.tex
new notebook; more text

\begin{figure}[t!]
    \begin{centering}
        \includegraphics[width=\linewidth]{figures/nullspace_ensemble.pdf}
        \oscaption{nullspace_ensemble}{%
            Posterior shrinkage as a function of spherical harmonic degree
            for a hypothetical scenario in which an observer can measure the light curves
            of 1 (blue), 3 (orange), 10 (green), and 30 (red) \emph{identical} stars
            viewed at random inclinations.
            The thicker curves correspond to
            the mean shrinkage for each experiment.
            The information content in the light curve of a star observed
            from a single vantage point approaches zero as $l$
            increases. However, observing many identical stars from different vantage points
            allows one to recover nearly all of the information in the
            even spherical harmonic modes. This is why ensemble analyses of
            many similar stars at different inclinations allows us to infer
            their surface properties.
            \label{fig:nullspace_ensemble}
        }
    \end{centering}
\end{figure}

@@

\begin{figure}[t!]
    \begin{centering}
        \includegraphics[width=\linewidth]{figures/evenmodes.pdf}
        \oscaption{evenmodes}{%
            A spotted stellar surface seen in Mollweide projection
            (left), the same surface with all odd $l > 2$ harmonics
            set to zero (center), and the same surface with $l = 1$
            and all even harmonics set to zero (right). The plot
            on the right represents the portion of the surface that is
            in the null space of the light curve problem at all
            inclinations; the plot in the center is the portion of the
            surface that can be inferred at some value of the inclination.
            \label{fig:evenmodes}
        }
    \end{centering}
\end{figure}

@@

\begin{figure}[t!]
    \begin{centering}
        \includegraphics[width=\linewidth]{figures/nullspace_ensemble_ld.pdf}
        \\
        \includegraphics[width=\linewidth]{figures/nullspace_ensemble_ld_var.pdf}
        \oscaption{nullspace_ensemble_ld}{%
            \emph{Top:} Same as Figure~\ref{fig:nullspace_ensemble},
            but for limb-darkened
            stars with coefficients $u_1 = 0.5$ and $u_2 = 0.25$. Odd
            modes can now be probed, at the expense of the even modes.
            \emph{Bottom:} Same as the top panel, but allowing for
            a 10\% variation in $u_1$ and $u_2$
            across different observations. For $\gtrsim 30$ light curves,
            there is virtually no null space up to $l_\mathrm{max} = 15$.
            \codelink{nullspace_ensemble_ld_var}
            \label{fig:nullspace_ensemble_ld}
        }
    \end{centering}
\end{figure}

@@

@@

\subsection{Time evolution}
\label{sec:temporal}
\xxx{TODO: The model presented in this paper assumes statics stellar surfaces,
    corresponding to perfectly periodic light curves. This is hardly ever the
    case in practice, as spots evolve and differential rotation changes the surface
    over time. We can still use this model, but we need to either restrict our analysis
    to segments of light curves of duration shorter than the timescale on which
    the surface evolves, or explicitly model the effect of time evolution on the covariance.
    The next paper will deal with temporal evolution more
    rigorously, but the way to do this is just take the elementwise product of the \starryprocess
    flux covariance with a temporal covariance. This corresponds to the Kronecker product
    in spherical harmonic space. Show a figure, give an example, etc.}

@@

\subsection{Gaussians all the way down}
\label{sec:other-marg}

@@

Given the complexity of the operations involved,
performing these marginalizations is beyond the scope of the present paper.
However, given the significant computational benefits of this marginalization,
and given the fact that the mapping problem is particularly sensitive to
the limb darkening coefficients (see \S\ref{sec:limbdark}), marginalizing over
$P$ and $\mathbf{u}$ will be the subject of the next paper in this series.

@@

In ensemble analyses, we must compute the likelihood of each of the $M$ light curves
conditioned on $\pmb{\theta}_\bullet$. In practice, each star will have a
different rotation period, different limb darkening coefficients, and different
photometric uncertainty, meaning we must factorize $M$ different covariance
matrices. Fortunately, the spherical harmonic covariance, $\pmb{\Sigma}_\mathbf{y}$
(\S\ref{sec:gp}) need only be computed once; this can then be linearly
transformed into the flux basis for each light curve. As the evaluation of
$\pmb{\Sigma}_\mathbf{y}$ is the computationally-intensive step, the
likelihood evaluation typically scales sub-linearly with $M$. Furthermore,
as we discussed in \S\ref{sec:other-marg}, it is possible to marginalize over
the period and limb darkening coefficients, which would remove the scaling
with $M$ entirely if the photometric precision were the same for all
light curves. Even if it is not, the algorithm can still be greatly sped up,
although an implementation of this is deferred to the next paper.
\xxx{@DFM: Is this true? I.e., is there an efficient way to solve Equation~(\ref{eq:log-like})
    when $\pmb{\Sigma}$ is the same for all stars but $\mathbf{C}$ is different? Is there
    some limit in which an approximation of some sort can be made?}

>>>>

\begin{figure}[t!]
    \begin{centering}
        \includegraphics[width=\linewidth]{figures/nullspace_ensemble.pdf}
        \oscaption{nullspace_ensemble}{%
            Posterior shrinkage as a function of spherical harmonic degree
            for a hypothetical scenario in which an observer can measure the light curves
            of 1 (blue), 3 (orange), 10 (green), and 30 (red) \emph{identical} stars
            viewed at random orientations.
            The thicker curves correspond to
            the mean shrinkage for each experiment.
            The information content in the light curve of a star observed
            from a single vantage point approaches zero as $l$
            increases. However, observing many identical stars from different vantage points
            allows one to recover nearly all of the information in the
            even spherical harmonic modes. This is why ensemble analyses of
            many similar stars at different inclinations allows us to infer
            their surface properties.
            \label{fig:nullspace_ensemble}
        }
    \end{centering}
\end{figure}

@@

\begin{figure}[t!]
    \begin{centering}
        \includegraphics[width=\linewidth]{figures/evenmodes.pdf}
        \oscaption{evenmodes}{%
            A spotted stellar surface seen in Mollweide projection
            (left), the same surface with all odd $l > 2$ harmonics
            set to zero (center), and the same surface with $l = 1$
            and all even harmonics set to zero (right). The plot
            on the right represents the portion of the surface that is
            in the null space of the light curve problem at all
            inclinations; the plot in the center is the portion of the
            surface that can be inferred by combining observations
            made from all possible inclinations.
            \label{fig:evenmodes}
        }
    \end{centering}
\end{figure}

@@

\begin{figure}[t!]
    \begin{centering}
        \includegraphics[width=\linewidth]{figures/nullspace_ensemble_ld.pdf}
        \\
        \includegraphics[width=\linewidth]{figures/nullspace_ensemble_ld_var.pdf}
        \oscaption{nullspace_ensemble_ld}{%
            \emph{Top:} Same as Figure~\ref{fig:nullspace_ensemble},
            but for limb-darkened
            stars with quadratic coefficients $u_1 = 0.5$ and $u_2 = 0.25$. Odd
            modes can now be probed, at the expense of the even modes.
            \emph{Bottom:} Same as the top panel, but allowing for
            a 10\% variation in $u_1$ and $u_2$
            across different observations. For $\gtrsim 30$ light curves,
            there is virtually no null space up to at least $l_\mathrm{max} = 15$.
            \codelink{nullspace_ensemble_ld_var}
            \label{fig:nullspace_ensemble_ld}
        }
    \end{centering}
\end{figure}

@@

\section{Extensions}
\label{sec:extensions}

@@

\subsection{Time evolution}
\label{sec:temporal}

Another limitation of the base algorithm...

\xxx{TODO: The model presented in this paper assumes statics stellar surfaces,
    corresponding to perfectly periodic light curves. This is hardly ever the
    case in practice, as spots evolve and differential rotation changes the surface
    over time. We can still use this model, but we need to either restrict our analysis
    to segments of light curves of duration shorter than the timescale on which
    the surface evolves, or explicitly model the effect of time evolution on the covariance.
    The next paper will deal with temporal evolution more
    rigorously, but the way to do this is just take the elementwise product of the \starryprocess
    flux covariance with a temporal covariance. This corresponds to the Kronecker product
    in spherical harmonic space. Show a figure, give an example, etc.}

@@

\subsection{Marginalizing over period and limb darkening}
\label{sec:other-marg}

@@

Given the complexity of the operations involved,
performing these marginalizations is beyond the scope of the present paper.
However, given the significant computational benefits of this marginalization,
and given the fact that the mapping problem is particularly sensitive to
the limb darkening coefficients (see \S\ref{sec:limbdark}), marginalizing over
$P$ and $\mathbf{u}$ will be the subject of an upcoming paper.

@@

In ensemble analyses, we must compute the likelihood of each of the $M$ light curves
conditioned on $\pmb{\theta}_\bullet$. In practice, each star will have a
different rotation period, different limb darkening coefficients, and different
photometric uncertainty, meaning we must factorize $M$ different covariance
matrices. Fortunately, the spherical harmonic covariance, $\pmb{\Sigma}_\mathbf{y}$
(\S\ref{sec:gp}) need only be computed once; this can then be linearly
transformed into the flux basis for each light curve. As the evaluation of
$\pmb{\Sigma}_\mathbf{y}$ is the computationally-intensive step, the
likelihood evaluation typically scales sub-linearly with $M$. Furthermore,
as we discussed in \S\ref{sec:other-marg}, it is possible to marginalize over
the period and limb darkening coefficients, which would remove the scaling
with $M$ entirely if the photometric precision were the same for all
light curves. Even if it is not, the algorithm can still be greatly sped up,
although an implementation of this is deferred to the next paper.
